# Getting Data into R

```{r installation-02, echo=T, message=FALSE, warning=FALSE, results='hide'}
install.packages("dplyr",repos = "https://cran.us.r-project.org")
install.packages("tidyr",repos = "https://cran.us.r-project.org")
install.packages("stringr",repos = "https://cran.us.r-project.org")
install.packages("lubridate",repos = "https://cran.us.r-project.org")
install.packages("readr",repos = "https://cran.us.r-project.org")
```

```{r load-libs-02, echo=T, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
```


Read in the data 
```{r read-data-02}
urlfile02a="https://raw.githubusercontent.com/apicellap/data/main/compensation.csv"
compensation<-read.csv(url(urlfile02a))
head(compensation)
```
## Checking that your data are your data

Generate names of the columns/variables in the console: 
```{r df-names}
names(compensation)
```

Produce number of observations (rows in each column) followed by # of variables:
```{r df-dimensions}
dim(compensation) 
```

Review structure of the data: 
```{r df-structure-02}
str(compensation)
```

## Appendix advanced activity: dealing with untidy data

```{r read-data-02b}
urlfile02b="https://raw.githubusercontent.com/apicellap/data/main/nasty%20format.csv"
nasty.format<-read.csv(url(urlfile02b))
head(nasty.format)
```

Review data structure: 
```{r df-structure-02b}
str(nasty.format)
```

- this dataset is poorly constructed 

Eliminate extra (37th) row in dataset: 
```{r elim-row}
nasty.format<-filter(nasty.format, Bottle !="") # '!=' symbol means '≠' 
tail(nasty.format)


```

- this filter function is programmed to capture every row in which variable, 'Bottle', contains text 

Create new variables and assort data into them:  
```{r gather-data}
tidy_data <- gather(nasty.format, 
                    Date, Abundance, #the variables to be created 
                    4:11) #column headers that are dates in the nasty.format dataframe 
head(tidy_data)

```

Remove the 'X', which precedes that date in each observation: 
```{r edit-dates}
tidy_data <- mutate(tidy_data, Date=substr(Date,2,20))
head(tidy_data)
```

Display all unique dates: 
```{r unique-dates}
unique(
  tidy_data$Date) #this says use the observations in the variable 'Date' in the 'tidy_data' dataframe
```

Reformat the dates to be universally recognized: 
```{r reformat-dates}
tidy_data <-mutate(tidy_data, Date=dmy(Date))
head(tidy_data)
```

`separate()` Separates information present in one column to multiple new columns 

`unite()` Puts information from several columns into one column 

`rbind()` Puts datasets with exactly the same columns together

`cbind()` Combines two datasets with exactly the same columns together

`full_join()` Joins two datasets with one or more columns in common 

`merge()` Same function as full_join() but from the base package

