# Data Management and Manipulation 

```{r install-03, echo=T, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
install.packages("dplyr",repos = "https://cran.us.r-project.org")
install.packages("tidyr",repos = "https://cran.us.r-project.org")
install.packages("stringr",repos = "https://cran.us.r-project.org")
install.packages("lubridate",repos = "https://cran.us.r-project.org")

```

```{r load-libs-03, echo=T, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)

```

Read in the data 
```{r read-data-03}
urlfile03a="https://raw.githubusercontent.com/apicellap/data/main/compensation.csv"
compensation<-read.csv(url(urlfile03a))
head(compensation)
```

Summarize the data in each variable:
```{r summary-data-03}
 summary(compensation)
```

## Subsetting 
Create new dataframe comprised of specific variable(s):
```{r subset-03a}
head(select(compensation,
           Fruit))
```

Select all columns except one: 
```{r select-03}
head(select(compensation, -Root))
```

Create new dataframe comrpised of specific variable(s) *except* 'Root': 
```{r slice-03}
head(slice(compensation, 2:10))
```

Create new dataframe comprised of a list of variables:
```{r slice-03b}
head(slice(compensation, c(2,3,10)))
```

Filter data set to only observations in which this is TRUE: 
```{r filter-03}
filter(compensation, Fruit == 80) 
```

Grab observations when Fruit is not equal to 80: 
```{r filter-03b}
head(filter(compensation, Fruit !=80))
```

Grab any observations in which Fruit is ≤ 80; can also use < symbol for less than:  
```{r filter-03c}
head(filter(compensation, Fruit <=80))
```

Grab any observations in which Fruit is greater than 95 OR less than 15: 
```{r filter-03d}
head(filter(compensation, Fruit >95|Fruit<15))
```

Grab any observations in which Fruit is greater than 50 AND less than 55: 
```{r filter-03e}
head(filter(compensation, Fruit >50 & Fruit<55))
```

Order data by Fruit from lowest to highest observation: 
```{r arrange-03}
head(arrange(compensation, Fruit))

```

Create new dataframe that filters observations that have Fruit values above 80 
and only contains the corresponding Root values: 
```{r filter-03-special}
head(select(filter(compensation, Fruit>80), Root))
```

## Calculating summary statistics about groups of your data 

Perform summary analyses on dataframe: 
```{r summary-stats-03}
summarise(
  group_by(compensation, Grazing), #access the dataframe, target Grazing to be the grouping variable 
  meanFruit = mean(Fruit)) #creates the object, meanFruit which is the mean of the data in the Fruit variable

```

Additional summary functions and create new dataframe to encompass calculations:
```{r summary-stats-df-03}
mean.fruit<-summarise(
  group_by(compensation, Grazing), 
  meanFruit = mean(Fruit), sdfruit =sd(Fruit)) #multiple statistics can be calculated within summarise 
mean.fruit
```


```{r special-func-03}
x <- sum(with(compensation, Grazing == "Grazed")) #counts number of observations for variable when it = Grazed 
x
```

```{r summary-stats-df-03b}

SE.mean.fruit<-summarise(
  group_by(compensation, Grazing), 
  meanFruit = mean(Fruit), 
  SEfruit =(sd(Fruit))/sqrt(x)) #multiple statistics can be calculated within summarise 
SE.mean.fruit

```

