# Visualizing your data

## The first step in every data analysis - making a picture

```{r install-04, echo=T, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
install.packages("ggplot2",  repos = "https://cran.us.r-project.org")
install.packages("dplyr",  repos = "https://cran.us.r-project.org")
```

```{r load-libs-04, echo=T, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(dplyr)
```

Read in the data 
```{r read-data-04}
urlfile04a="https://raw.githubusercontent.com/apicellap/data/main/compensation.csv"
compensation<-read.csv(url(urlfile04a))
```

View dataframe & read the variables + the first few of their observations horizontally: 
```{r glimpse-04}
glimpse(compensation) 
```

## ggplot2: a grammar for graphics 

Create base plot: 
```{r base-pointplot-04}
base_plot <-ggplot(compensation, aes(x = Root, y = Fruit, 
                                colour=Grazing)) + #colour: for the two levels of the categorical variable, Grazing
  geom_point()                                       
base_plot
```

Render background white instead of gray: 
```{r sp-mod1}
base_plot + theme_bw()
```

```{r sp-mod2}
base_plot + 
  theme_bw() + 
  geom_point(
    size = 5) #alter size of datapoints in scatterplot 
```

Add x and y axis titles: 
```{r sp-mod3}
base_plot + theme_bw() + geom_point(size = 5) + 
    xlab("Root Biomass") + 
  ylab("Fruit Production") 
```

## Box and whisker plots
```{r base-boxplot-04}
base_plot2 <- ggplot(compensation, aes(x = Grazing, y = Fruit)) + 
  geom_boxplot() +  
  geom_point(
    size = 4, #size of point
    colour = 'lightgrey', #color of point
    alpha = 0.5) +         #transparency of point 
  xlab("Grazing treatment") + 
  ylab("Fruit Production") + 
  theme_bw() 
base_plot2
```

## Distributions: making histograms of numeric variables 

```{r histo-04}
ggplot(compensation, aes(x=Fruit))+
  geom_histogram(bins=15)  #bins defines how many histogram bins there are 
 
```

```{r histo-04b}
ggplot(compensation, aes(x=Fruit))+
  geom_histogram(bins=15) + 
  facet_wrap(~Grazing) #facet_wrap() allows you to put the plots next to each other, a variable must be specified 
```


